Formal Verification of Real-Time Data Processing of the LHC Beam Loss Monitoring System: A Case Study

### 2011-03-30 fmics/paper.tex
merge changes to paper.tex from dropbox

We describe a collaborative effort in which the HOL4 theorem prover is being used to formally verify properties of a structure within the Large Hadron Collider (LHC) machine protection system at the European Organization for Nuclear Research (CERN).
This structure, known as \emph{Successive Running Sums} (SRS), generates the primary input to the decision logic that must initiate a critical action by the LHC machine protection system in response to the detection of a dangerous level of beam particle loss.
The use of mechanized logical deduction complements an intensive study of the SRS structure using simulation.
We are especially interested in using logical deduction to obtain a generic result that will be applicable to variants of the SRS structure.
This collaborative effort has individuals with diverse backgrounds ranging from theoretical physics to system safety.
%One interesting result is the extent to which
The use of a formal method has compelled the stakeholders to clarify intricate details of the SRS structure and behaviour.
\end{abstract}

@@

%************** Rushby's quote to be added later on *****
%Rushby's emphasis on the role of mechanized logical deduction as an "implacable skeptic that insists on its human user providing justification" thus rings true.

@@

The loss measurement principle \cite{Dehning-IPAC,Chris-FPGA} is based on the energy deposition detection of secondary shower particles using a specially-designed detector called \emph{ionization chamber}. There are around 4000 ionization chambers located under the magnets all around the tunnel. These detectors convert the signal given by the recording of shower particles into electrical signals and send them to tunnel cards. Tunnel cards, which are located in the tunnel and implemented by radiation-tolerant electronics, acquire and digitize the data from the detectors and transmit those to the surface using optical links. There, the data processing cards, named BLETCs, receive those data and decide whether or not the beam should be permitted to be injected or continue circulating. Each tunnel card receives data from eight detectors and each surface card receives data from two tunnel cards. The surface card provides data to the Logging, the Post Mortem and the Collimation systems that will be used to drive on-line displays in the control room to allow offline analysis of the losses and to setup automatically the collimators. Due to demanding performance requirements, the BLMS is implemented on modern FPGAs, which include the resources needed to design complex processing and can be reprogrammed making them ideal for future upgrades or system specification changes. Figure~\ref{fig:BLETC} shows a block diagram of the processes assigned in BLETC FPGA. In the following, we give a brief overview of each of the four main processing blocks in BLETC card.

@@

\begin{table}[t]
\centering
\begin{tabular}{|r|r|r|r|c|l|}
\hline
\multicolumn{2}{|c}{Range} &\multicolumn{2}{|c|}{Refreshing} & & \\\cline{1-4}
40 {\textmu}s steps & ms & 40 {\textmu}s steps  & ms & slice & running sum \\\hline\hline
1 & 0.04 & 1 & 0.04 & & RS0 \\\hline
2 & 0.08 & 1 & 0.04 & & RS1 \\\hline
8 & 0.32 & 1 & 0.04 & slice 1 & RS2 \\\hline
16 & 0.64 & 1 & 0.04 & & RS3 \\\hline
64 & 2.56 & 2 & 0.08 & slice 2 & RS4 \\\hline
256 & 10.24 & 2 & 0.08 & & RS5 \\\hline
2048 & 81.92 & 64 & 2.56 &slice  3 & RS6 \\\hline
16384 & 655.36 & 64 & 2.56 &slice 3 & RS7 \\\hline
32768 &1310.72 & 2048 & 81.92 & slice 4 & RS8 \\\hline
131072 & 5242.88 & 2048 & 81.92 & slice 4 & RS9 \\\hline
524288 & 2097.52 & 16384 & 655.36 & slice 5 &RS10 \\\hline
2097152 & 83886.08 & 16384 & 655.36 & slice 5 & RS11 \\\hline
\end{tabular}
\vspace{0.04in}
\caption{{\bf<missing caption>}
\label{fig:example-bakery}}
\end{table}

@@

Figure~\ref{fig:auxdefs} defines a set of additional functions required to prove our main results.
The function $\mathsf{delay\_sum}\;n$ represents the cumulative delay of the preceding shift registers of a slice.
For example, $\mathsf{delay\_sum}\;4 = \mathsf{delay}\;3 + \mathsf{delay}\;1 + \mathsf{delay}\;0 = 2 +1 +1 = 4$.
The function $\mathsf{last\_update}\;n\;t$ specifies the time that the $n$-th slice last updated, and $\mathsf{exact}\;D\;n\;x\;t$ computes the exact sum of consecutive input counts, without delay, that $\mathsf{output}\;D\;n\;x\;t$ is supposed to approximate.
% (so $\mathsf{exact}$ is to $\mathsf{true\_sum}$ as $\mathsf{output}$ is to $\mathsf{RS}$).

@@

\begin{proof}
The shift register of slice $n$ is updated every $\mathsf{delay}\;n$ time steps.
When updated, the values in its cells are shifted one cell.
%The shift register cells get their values from the previous cell but they only update every $\mathsf{delay}\;n$ time steps. Therefore
Therefore\footnote{For presentation convenience, here, we do not provide the details regarding the case when $t$ is small, for example $t<m\times\mathsf{delay}\;n$.
For complete theorem statements and the proof script, please refer to \url{https://github.com/xrchz/CERN-LHC-BLMTC-SRS/blob/master/hol/srsScript.sml}.},
\begin{equation*}
\mathsf{SR}\;D\;n\;(m+1)\;t=\mathsf{SR}\;D\;n\;m\;(t-\mathsf{delay}\;n)
\end{equation*}
By induction on $m$, using the above,
\begin{equation*}
\mathsf{SR}\;D\;n\;m\;t=\mathsf{SR}\;D\;n\;0\;(t-(m\times\mathsf{delay}\;n))
\end{equation*}
By induction on $t$, we can show that $\mathsf{output}\;D\;n\;x\;t$ is a sum of values of consecutive shift register cells
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\mathsf{SR}\;D\;n\;m\;t
\end{equation*}
Combining the last two results, $\mathsf{output}\;D\;n\;x\;t$ is a sum of consecutive values of the first shift register cell, $\mathsf{SR}\;D\;n\;0$.
Thus, we can express $\mathsf{output}\;D\;n\;x\;t$ in terms of $\mathsf{output}\;D\;n'\;x'\;t$, where $(n',x')=\mathsf{input}\;n$, since the source of $\mathsf{SR}\;D\;n\;0$ is the previous (input) slice's output.
Finally, by induction on $n$, we can express $\mathsf{output}\;D\;n\;x\;t$ as a function of $D$ alone, as required, using the result above for the inductive case.
\qed
\end{proof}

@@

We formulate the difference between the $\mathsf{RS}$ values and their corresponding $\mathsf{true\_sum}$s, caused by the delay, as $\mathsf{error}$. 
In order to define a maximum bound on the $\mathsf{error}$, we had to formulate some characterization of the input stream to the SRS structure as constraints on $D$.
One of these constraints is a bound on the difference between two consecutive count values.

@@

%One of our objectives in formally modelling SRS was to keep the topology parameterized.
%It is likely that the configuration of the slices and shift registers will in the future change to accommodate higher precision.
%A parameterized model can be used both for current and future versions of the system.
%For this, we formalized the definitions of the model by six parameters:

%Our work proceeded in two phases, starting with a model of a simplified structure and proceeding to a model which is closer to the SRS implementation in BLMS.
%In both phases, we model each building block of the SRS structure that holds a count value -- for example, each cell in each shift register -- as a function in HOL that takes the input stream and current time step as inputs and returns the value the corresponding building block holds at that time.

%Our simplifications in the first phase include:
%\begin{enumerate}
%\item Masking the middle taps in each slice,
%\item Considering a more regular arrangement of slices than the SRS implementation in BLMS,
%\item Masking the subtractors and the accumulators in each slice, instead, defining the output of a slice as the sum of the values in its cells directly,
%\item Simplifying timing by ignoring some of the delays that each building block may introduce, {\bf this is not true - there were delays in the simplified model}
%\item Using natural numbers, rather than finite words, for count values.
%\end{enumerate}

%Building a logical model for simplified structure of the SRS helped us improve our understanding of the behavior of the SRS.

%In the second phase, we eliminated the first four simplifications (listed above) by adding middle taps, a more flexible arrangement of slices, the subtractor and accumulator, and considering the delays that each block may introduce.
%In both cases, however, the model is simultaneously a simplification and a generalization of the actual structure of SRS in BLMS in the following way: we model an infinite number of slices (rather than six slices), each with an infinite number of shift register cells and taps (rather than fixed width shift registers and only two taps), simply by letting indices range over the natural numbers without explicitly giving limits.
%This parameterization makes the model and the analysis more likely to be applicable to future upgraded/modified versions of the system.

@@

>>>>

We describe a collaborative effort in which the HOL4 theorem prover is being used to formally verify properties of a structure within the Large Hadron Collider (LHC) machine protection system at the European Organization for Nuclear Research (CERN).
This structure, known as \emph{Successive Running Sums} (SRS), generates the primary input to the decision logic that must initiate a critical action by the LHC machine protection system in response to the detection of a dangerous level of beam particle loss.
The use of mechanized logical deduction complements an intensive study of the SRS structure using simulation.
We are especially interested in using logical deduction to obtain a generic result that will be applicable to variants of the SRS structure.
This collaborative effort has individuals with diverse backgrounds ranging from theoretical physics to system safety.
The use of a formal method has compelled the stakeholders to clarify intricate details of the SRS structure and behaviour.
\end{abstract}

@@

@@

The loss measurement principle \cite{Dehning-IPAC,Chris-FPGA} is based on the energy deposition detection of secondary shower particles using a specially-designed detector called \emph{ionization chamber}.
There are around 4000 ionization chambers strategically placed on the sides of the magnets all around the tunnel.
These detectors convert the signal given by the recording of shower particles into electrical signals and send them to tunnel cards.
Tunnel cards, which are located in the tunnel and implemented by radiation-tolerant electronics, acquire and digitize the data from the detectors and transmit those to the surface using optical links.
There, the data processing cards, named BLETCs, receive those data and decide whether or not the beam should be permitted to be injected or continue circulating.
Each tunnel card receives data from eight detectors and each surface card receives data from two tunnel cards.
The surface card provides data to the Logging, the Post Mortem and the Collimation systems that will be used to drive on-line displays in the control room, perform long-term storage for offline analysis, and to setup the collimators automatically.
Due to demanding performance requirements, the BLMS is implemented on modern FPGAs, which include the resources needed to design complex processing and can be reprogrammed making them ideal for future upgrades or system specification changes.
Figure~\ref{fig:BLETC} shows a block diagram of the processes assigned in BLETC FPGA.
In the following, we give a brief overview of each of the four main processing blocks in BLETC card.

@@

\begin{table}[t]
\centering
\begin{tabular}{|r|r|r|r|c|l|}
\hline
\multicolumn{2}{|c}{Range} &\multicolumn{2}{|c|}{Refreshing} & & \\\cline{1-4}
40 {\textmu}s steps & ms & 40 {\textmu}s steps  & ms & slice & running sum \\\hline\hline
1 & 0.04 & 1 & 0.04 & slice 1 & RS0 \\\hline
2 & 0.08 & 1 & 0.04 & slice 1 & RS1 \\\hline
8 & 0.32 & 1 & 0.04 & slice 2 & RS2 \\\hline
16 & 0.64 & 1 & 0.04 & slice 2 & RS3 \\\hline
64 & 2.56 & 2 & 0.08 & slice 3 & RS4 \\\hline
256 & 10.24 & 2 & 0.08 & slice 3 & RS5 \\\hline
2048 & 81.92 & 64 & 2.56 &slice 4 & RS6 \\\hline
16384 & 655.36 & 64 & 2.56 &slice 4 & RS7 \\\hline
32768 &1310.72 & 2048 & 81.92 & slice 5 & RS8 \\\hline
131072 & 5242.88 & 2048 & 81.92 & slice 5 & RS9 \\\hline
524288 & 2097.52 & 16384 & 655.36 & slice 6 &RS10 \\\hline
2097152 & 83886.08 & 16384 & 655.36 & slice 6 & RS11 \\\hline
\end{tabular}
\vspace{0.04in}
\caption{{\bf<missing caption>}
\label{fig:example-bakery}}
\end{table}

@@

Figure~\ref{fig:auxdefs} defines a set of additional functions required to prove our main results.
The function $\mathsf{delay\_sum}\;n$ represents the cumulative delay of the preceding shift registers of a slice.
For example, $\mathsf{delay\_sum}\;4 = \mathsf{delay}\;3 + \mathsf{delay}\;1 + \mathsf{delay}\;0 = 2 +1 +1 = 4$.
The function $\mathsf{last\_update}\;n\;t$ specifies the time that the $n$-th slice last updated, and $\mathsf{exact}\;D\;n\;x\;t$ computes the exact sum of consecutive input counts, without delay, that $\mathsf{output}\;D\;n\;x\;t$ is supposed to approximate.

@@

\begin{proof}
The shift register of slice $n$ is updated every $\mathsf{delay}\;n$ time steps.
When updated, the values in its cells are shifted one cell.
Therefore\footnote{For presentation convenience, here, we do not provide the details regarding the case when $t$ is small, for example $t<m\times\mathsf{delay}\;n$.
For complete theorem statements and the proof script, please refer to \url{https://github.com/xrchz/CERN-LHC-BLMTC-SRS/blob/master/hol/srsScript.sml}.},
\begin{equation*}
\mathsf{SR}\;D\;n\;(m+1)\;t=\mathsf{SR}\;D\;n\;m\;(t-\mathsf{delay}\;n)
\end{equation*}
By induction on $m$, using the above,
\begin{equation*}
\mathsf{SR}\;D\;n\;m\;t=\mathsf{SR}\;D\;n\;0\;(t-(m\times\mathsf{delay}\;n))
\end{equation*}
By induction on $t$, we can show that $\mathsf{output}\;D\;n\;x\;t$ is a sum of values of consecutive shift register cells
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\mathsf{SR}\;D\;n\;m\;t
\end{equation*}
Combining the last two results, $\mathsf{output}\;D\;n\;x\;t$ is a sum of consecutive values of the first shift register cell, $\mathsf{SR}\;D\;n\;0$.
Thus, we can express $\mathsf{output}\;D\;n\;x\;t$ in terms of $\mathsf{output}\;D\;n'\;x'\;t$, where $(n',x')=\mathsf{input}\;n$, since the source of $\mathsf{SR}\;D\;n\;0$ is the previous (input) slice's output.
Finally, by induction on $n$, we can express $\mathsf{output}\;D\;n\;x\;t$ as a function of $D$ alone, as required, using the result above for the inductive case.
\qed
\end{proof}

@@

We formulate the difference between the $\mathsf{RS}$ values and their corresponding $\mathsf{true\_sum}$s, caused by the delay, as $\mathsf{error}$.
In order to define a maximum bound on the $\mathsf{error}$, we had to formulate some characterization of the input stream to the SRS structure as constraints on $D$.
One of these constraints is a bound on the difference between two consecutive count values.
Such a bound can be derived from the highest quench level thresholds.

@@

@@

We described a case study in which HOL4 theorem prover is used to formally verify properties of the SRS structure within the LHC machine protection system.
In this case study, we built a parameterized model of the SRS structure and showed that its behavior is correct with respect to its specification.
It is likely that the configuration of the slices and shift registers will change in the future to accommodate higher precision.
Thus, the parameterization in our model was crucial to make it applicable to the future upgrades.

One of our main challenges in this effort was building the formal model and formulating the correctness statements.
To understand the structure of the SRS, we started with a model of a simplified structure, which had a more regular arrangements of the slices and ignored the middle taps, and proceeded to a model which is closer to the SRS implementation in BLMS.
In addition, we used a basic spreadsheet simulation model for a few slices for sanity testing the correctness formulae.
However, some sanity tests related to Theorem 1 led us astray when we did not know the correct formula for that theorem.
Because the formulae we conjectured worked for small values of $n$, and simulating large values of $n$ was expensive, but the counterexamples were only on large values of $n$.
The most important use of mechanized logical deduction in this effort was its role as an ``implacable skeptic" that insisted on us providing the justification and compelled us to clarify the intricate details of the SRS structure and behavior.

The use of mechanized theorem proving to verify the correctness of the SRS behavior complements an intensive verification based on simulation performed at CERN.
The simulation was mainly based on two scenarios one for recording fast losses and  another for the situation when a loss gradually builds up in the accelerator.
The results of the simulation showed that the current implementation of SRS, given in Figure~\ref{fig:srs}, satisfies the specification, that is the running sums have  the maximum 20\% relative error.
On the contrary, the results in this paper apply to all possible input steams, not just a limited number of input streams simulating fast or slow loss scenarios.
But, the error bound given in Theorem~\ref{max_error} does not satisfy the maximum 20\% relative error.
In order to make this bound tighter, we need to provide more characterization of the input as constraints on $D$.
Due to the nature of the problem, such a characterization of input is not trivial.
Therefore, the SRS structure in LHC is an interesting example in which the formal verification does not completely replace conventional verification approaches such as simulation.

%Future direction, go with lower level implementation rather than finite words (closer  to vhdl coed) and improve the bound
